Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters

### 2019-05-18 paper.tex

\documentclass{IEEEtran}
\usepackage[hyphens]{url}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{colortbl} % for \rowcolor
\usepackage{ulem} % for \uwave
\usepackage{caption}
\usepackage{subcaption}
\usepackage[bookmarks=false]{hyperref}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{algorithm}
\usepackage{algorithmicx} % Doc is at http://tug.ctan.org/macros/latex/contrib/algorithmicx/algorithmicx.pdf
\usepackage{algpseudocode}
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algblock{Input}{EndInput}
\algnotext{EndInput}
\algblock{Output}{EndOutput}
\algnotext{EndOutput}
\newcommand{\Desc}[2]{\State \makebox[2em][l]{#1}#2}

@@

\title{Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters}
\author{
    \IEEEauthorblockN{
        Val\'erie Hayot-Sasson and Tristan Glatard \\
        Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada
    }
    \IEEEauthorblockA{}
}
\maketitle

@@

% Goal of the paper
In this paper, we study the use of pilot jobs to run scientific
applications with Apache Spark~\cite{zaharia2016apache} on shared HPC
clusters. The method currently recommended for this purpose involves batch
requesting all the
necessary resources and launching a standalone Spark cluster once the
resources have been allocated. This is for instance the method used in
Compute Canada (see details \href{https://docs.computecanada.ca/wiki/Apache_Spark/en}{here}),
 our national computing infrastructure. With pilot
jobs, rather than requesting all the resources at once, a Spark cluster is
launched with a subset of the resources and is expanded as more resources
get allocated. Consistently with previous examples of pilot job
deployments, we hypothesize that this strategy would reduce queuing times
by (1) fragmenting resource requirements, and (2) allowing shorter walltime
estimates. While there have been some efforts on implementing pilot jobs
for Apache Spark~\cite{jha-spark-pbs}, research is limited and
none of the them detail their quantitative effect.

@@

To summarize, our paper
makes the following contributions:
\begin{itemize}
\item We present a lightweight pilot-job framework to run Apache Spark
applications on HPC clusters.
\item We compare the makespan of a typical neuroimaging Big Data
application with and without pilot jobs on two different HPC clusters and
in different conditions.
\item We describe a simple performance model to validate that the observed
differences come from variations in queuing times rather than other
factors.
\end{itemize}
Our methods, including infrastructure, job templates, application and
performance model are described in Section~\ref{sec:methods}.
Section~\ref{sec:results} presents our results which are discussed in
Section~\ref{sec:discussion}. Section~\ref{sec:conclusion} concludes on the 
relevant of pilot jobs for Apache Spark applications on HPC clusters.

@@

    \TG{A sub-section could be added here to describe the main features of Spark used:
    client mode, cluster mode, master selection, supervise mode, etc}

    \section{Spark Configuration}
    Spark has many configuration options available for dynamic resource provisioning.
    Its standalone cluster provides two deployment modes: client and cluster. The client
    deploy mode executes the driver within the \texttt{spark-submit} process as
    a client to the cluster. The cluster deploy-mode, however, runs the driver within
    a worker process. Such a configuration is practical when the driver program
    is deployed on a machine not co-located in the cluster to reduce network 
    latency between workers and driver. The cluster deploy mode also has the texttt{supervise}
    feature which provides fault-tolerance to the driver. This is particularly useful for
    pilot-based overlay clusters where the running driver may fail to walltime expiration.
    Cluster deploy mode is not available for python applications using the standalone mode.

    With HPC clusters, client mode is not necessarily possible on the login node as the
    driver client may require more resources than permitted. Moreover, due to cluster
    security policies, it may not be possible, or difficult to execute the driver from
    an external computer, such as a personal laptop. Therefore, for both batch and pilot
    scheduling, it is necessary to launch the driver from within the Slurm jobs.

    Spark also provides master fault tolerance through single-node recovery, where a shared
    filesystem is available. To enable this, a recovery folder must be set. Should the master
    fail, a new master will be able to takeover based on the information available in the recovery
    folder. Although we start a master on each pilot, we have not incorporated this recovery function
    for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
    were any nodes can fail by design.

@@

@@

    The pilot submission template is similar to that of the batch template, however,
    each pilot will start its own Spark master and worker. Only one pilot,
    the first pilot to attempt to launch the driver, will do so. The reason for 
    which each pilot starts its own master is to ensure the fault tolerance of the 
    masters. In this sort of configuration, should the active master be killed, one
    of the stand-by masters will takeover and the application will be able to 
    resume. Such a configuration is particularly favourable in pilot scheduling 
    scenarios as node failures may be more frequent due to walltime expiration.
    Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
    This would not only allow the driver to be executed directly on one of the workers,
    but also allow us to make the driver fault-tolerant through the
    \texttt{supervise} mode, which is only available in cluster deploy. As
    with the masters, it is particularly important to have a fault-tolerant
    driver in pilot-scheduling scenarios due to possible walltime
    expiration.
    Should pilots be idle for a certain duration, the
    pilots will terminate themselves such as to not hog resources.
    % \TG{It should be explained before that one of the interests
    % of pilots is to ``play'' with shorter walltimes}. \TG{It would also be
    % nice to have runs where pilots 
    % have shorter walltimes, to see if queuing times are shorter too.}

@@

    \subsection{Application}
        \begin{algorithm}\caption{Incrementation}\label{alg:incrementation}              
            \begin{algorithmic}[1]                                                       
                \Input                                                                       
                    \Desc{$x$}{a sleep delay in seconds}                                         
                    \Desc{$n$}{a number of iterations}                                           
                    \Desc{$C$}{a set of image chunks}                                            
                \EndInput                                                                    
                \ForEach{$chunk \in C$}                                                      
                    \State read $chunk$                                        
                    \For{$i \in [1, n]$}                                                         
                        \State $chunk\gets chunk+1$                                              
                        \State sleep $x$                               
                    \EndFor                                                                      
                    \State write $chunk$                                            
                \EndFor                                                                      
            \end{algorithmic}                                                                
        \end{algorithm}
    To determine the added value of pilot scheduling over batch scheduling of Spark
    applications, we required a Spark application operating on a large dataset with
    an important processing time to emulate what would be the average requirements
    of a scientific Spark application. For this, we created a synthetic application 
    that would process the Big Brain~\cite{amunts2013bigbrain}, a 76~G 3D histological
    image of a human brain. The algorithm is a chain of map transformations that
    at each transformation increment the voxels of the image by 1 (see Algorithm~\ref{alg:incrementation}).
    We chose such a synthetic algorithm
    as the focus of our experiments is pilot scheduling and not the application in
    itself. Furthermore, this algorithm enabled us to have control over the task duration
    which was representative of scientific applications. Additionally, it was important
    that the overall application duration did not vary between the different levels of
    parallelism within our experiments. Being able to adjust the task duration based on
    level of parallelism allowed us to achieve this.

@@

    Therefore, assuming a fixed total CPU time, the relation
    between batch and pilot jobs can be represented as:
    \begin{equation}
        \frac{M_{batch}}{M_{pilot}} = \frac{W_{pilot}}{W_{batch}}\label{eq:makespancomp}
    \end{equation}
    where:
    \begin{itemize}
        \item $M_{batch}$ is the makespan of the batch application
        \item $M_{pilot}$ is the makespan of the pilot application
        \item $W_{pilot}$ is the average number of workers of the pilot application
        \item $W_{batch}$ is the average number of workers of the batch application
    \end{itemize}
    We will use this relation to discuss our results later on. It
    corresponds to an ideal case where no data or other overhead is
    present: only queuing times are included.

@@

As can be seen in both Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar}, pilots 
generally did not bring any performance improvement to the execution of Spark workloads
on B\'eluga and Cedar. In B\'eluga (Figure~\ref{fig:beluga1}), 16 pilots were always 
slower than batch except for repetitions 4 and 12. The maximum speedup achieved
was approximately 1.3~x for 8 pilots and 1.1~x for 16 pilots. In Figures~\ref{fig:beluga2}
and ~\ref{fig:beluga3}, it is also the case that batch is generally faster than
pilots for the majority of the cases, except a couple, in which the speedup brought
by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
that we see a maximum speedup of 2.8, however, it generally remains that pilots are slower. \TG{could you report the 
average speed-up of pilots?}

@@

In order to determine whether queueing time was the cause of the difference 
in makespans between batch and pilots, we compared makespan with queuing time of 
first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
makespans is primiarily due to queueing time.

@@

In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
given an experiment configuration, where the line corresponds to the estimated makespan using 
our model. Makespan for the model was calculated as:
$$
M = \frac{125\times(d + 20)}{W}\times 10,
$$
where:
\begin{itemize}
    \item $125$ is the number of chunks BigBrain was split into
    \item $M$ is the makespan of the application
    \item $W$ is the average number of Spark workers throughout the execution
    \item $d$ is the task delay associated with a given configuration
    \item $20$ is the average measured incrementation duration, in seconds, for a BigBrain chunk
    \item $10$ is the number of iterations
\end{itemize}
As can be seen in the figure, although pilots and batch had similar makespans
for a given number of average workers, pilots were consistently slower than batch
for the same number of workers. Both batch and pilots were consistent with the 
model, however \TG{there is a contradiction here: if both batch and pilots were consistent with the model, then
they should have similar makespans.}. In general, both pilots and batch had the same number of average
workers. \TG{You could also plot the difference in the number of workers, for each experiment.}

@@

>>>>

\documentclass{IEEEtran}
\usepackage[hyphens]{url}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{colortbl} % for \rowcolor
\usepackage{ulem} % for \uwave
\usepackage{caption}
\usepackage{subcaption}
\usepackage[bookmarks=false]{hyperref}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{algorithm}
\usepackage{algorithmicx} % Doc is at http://tug.ctan.org/macros/latex/contrib/algorithmicx/algorithmicx.pdf
\usepackage{algpseudocode}
\usepackage{stmaryrd}
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algblock{Input}{EndInput}
\algnotext{EndInput}
\algblock{Output}{EndOutput}
\algnotext{EndOutput}
\newcommand{\Desc}[2]{\State \makebox[2em][l]{#1}#2}

@@

\title{Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters}
\author{
    \IEEEauthorblockN{
        Val\'erie Hayot-Sasson and Tristan Glatard \\
        Department of Computer Science and Software Engineering, Concordia University \\
        Montr\'eal, Qu\'ebec, Canada
    }
    \IEEEauthorblockA{}
}
\maketitle

@@

% Goal of the paper
In this paper, we study the use of pilot jobs to run scientific
applications with Apache Spark~\cite{zaharia2016apache} on shared HPC
clusters. The method currently recommended for this purpose involves batch
requesting all the
necessary resources and launching a standalone Spark cluster once the
resources have been allocated. This is for instance the method used in
Compute Canada (see details \href{https://docs.computecanada.ca/wiki/Apache_Spark/en}{here}),
 our national computing infrastructure. With pilot
jobs, rather than requesting all the resources at once, a Spark cluster is
launched with a subset of the resources, and is expanded as more resources
get allocated. Consistently with previous examples of pilot job
deployments, we hypothesize that this strategy would reduce queuing times
by (1) fragmenting resource requirements, and (2) allowing shorter walltime
estimates. While there have been some efforts on implementing pilot jobs
for Apache Spark~\cite{jha-spark-pbs}, research is limited and
none of the them detail their quantitative effect.

@@

To summarize, our paper
makes the following contributions:
\begin{itemize}
\item We present SPA, a lightweight pilot-job framework to run Apache Spark
applications on HPC clusters.
\item We compare the makespan of a typical neuroimaging Big Data
application with and without pilot jobs on two different HPC clusters and
in different conditions.
\item We describe a simple performance model to validate that the observed
differences come from variations in queuing times rather than other
factors.
\end{itemize}
Our methods, including infrastructure, job templates, application and
performance model are described in Section~\ref{sec:methods}.
Section~\ref{sec:results} presents our results which are discussed in
Section~\ref{sec:discussion}. Section~\ref{sec:conclusion} concludes on the 
relevant of pilot jobs for Apache Spark applications on HPC clusters.

@@

    It is important to note that these clusters were used in production,
    that is, concurrently with other users. This allowed us to test the
    added-value of pilot scheduling in different realistic conditions of queuing
    times. At the time of our experiments, B\'eluga had recently been
    commissioned, resulting in low usage and shorter queuing times overall.
    Conversely, Cedar had been operating for a few years, resulting in
    higher usage and longer queuing times. Our experiments were repeated 15
    times in each configuration to capture queuing time variability.

    \subsection{Spark Configuration}

    \tristan{add a paragraph to explain master, worker, driver}

    Spark has many configuration options available for dynamic resource provisioning.
    Its standalone cluster provides two deployment modes: client and cluster. The client
    deploy mode executes the driver within the \texttt{spark-submit} process as
    a client to the cluster. The cluster deploy-mode, however, runs the driver within
    a worker process. Such a configuration is practical when the driver program
    is deployed on a machine not co-located in the cluster to reduce network 
    latency between workers and driver. The cluster deploy mode also has the \texttt{supervise}
    feature which provides fault-tolerance to the driver \tristan{, by means of checkpointing?}.
    This is particularly useful for
    pilot-based overlay clusters where the running driver may fail due to walltime expiration.
    Cluster deploy mode is not available for Python applications using the standalone mode \tristan{standalone mode wasn't defined.
    Rewrite positively saying that it's ony available in Scala?}.

    With HPC clusters, client mode is not necessarily possible on the login node as the
    driver client may require more resources than permitted. Moreover, due to cluster
    security policies, it may not be possible, or difficult to execute the driver from
    an external computer, such as a personal laptop. Therefore, for both batch and pilot
    scheduling, it is necessary to launch the driver from within the Slurm jobs, that is, in cluster mode.

    Spark also provides master fault tolerance through single-node recovery, where a shared
    filesystem is available. To enable this, a recovery folder must be set. Should the master
    fail, a new master will be able to takeover based on the information available in the recovery
    folder. Although we start a master on each pilot, we have not incorporated this recovery function
    for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
    were any nodes can fail by design. \tristan{Last two sentences should be moved to job templates.}

@@

    \tristan{This section could be renamed to ``SPA system design'', and you could add a paragraph explaining
    that for the sake of the experiment, you reimplemented the simplest possible pilot system since existing ones (refs)
    were either not found online, or difficult to use.}

@@

    The pilot submission template is similar to that of the batch template, however,
    each pilot will start its own Spark master and worker. Only one pilot,
    the first pilot to attempt to launch the driver, will do so \tristan{will do what? launch the driver?}.
    The reason for 
    which each pilot starts its own master is to ensure the fault tolerance of the 
    masters. In this sort of configuration, should the active master be killed, one
    of the stand-by masters will takeover and the application will be able to 
    resume. Such a configuration is particularly favourable in pilot scheduling 
    scenarios as node failures may be more frequent due to walltime expiration.
    Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
    This would not only allow the driver to be executed directly on one of the workers,
    but also allow us to make the driver fault-tolerant through the
    \texttt{supervise} mode, which is only available in cluster deploy. As
    with the masters, it is particularly important to have a fault-tolerant
    driver in pilot-scheduling scenarios due to possible walltime
    expiration.
    Should pilots be idle for a certain duration, the
    pilots will terminate themselves such as to not hog resources.
    % \TG{It should be explained before that one of the interests
    % of pilots is to ``play'' with shorter walltimes}. \TG{It would also be
    % nice to have runs where pilots 
    % have shorter walltimes, to see if queuing times are shorter too.}
    \TG{You could add a figure showing the driver, workers, master and stand-by masters, to make the system 
    description clearer.}

@@

    \subsection{Application}
        \begin{algorithm}\caption{Incrementation}\label{alg:incrementation}              
            \begin{algorithmic}[1]                                                       
                \Input                                                                       
                    \Desc{$x$}{a sleep delay in seconds}                                         
                    \Desc{$n$}{a number of iterations}                                           
                    \Desc{$C$}{a set of image chunks}                                            
                \EndInput                                                                    
                \ForEach{$chunk$ in $C$}                                                      
                    \State read $chunk$                                        
                    \For{$i \in \llbracket 1, n \rrbracket$}                                                         
                        \State $chunk\gets chunk+1$                                              
                        \State sleep $x$                               
                    \EndFor                                                                      
                    \State write $chunk$                                            
                \EndFor                                                                      
            \end{algorithmic}                                                                
        \end{algorithm}
    To determine the added value of pilot scheduling over batch scheduling of Spark
    applications, we required a Spark application operating on a large dataset with
    an important processing time to emulate what would be the average requirements
    of a scientific Spark application. For this, we created a synthetic application 
    that would process the Big Brain~\cite{amunts2013bigbrain}, a 76~G 3D histological
    image of a human brain. The algorithm is a chain of map transformations that
    at each transformation increment the voxels of the image by 1 (see Algorithm~\ref{alg:incrementation}).
    We chose such a synthetic algorithm
    as the focus of our experiments is pilot scheduling and not the application in
    itself. Furthermore, this algorithm enabled us to have control over the task duration
    which was representative of scientific applications. Additionally, it was important
    that the overall application duration did not vary between the different levels of
    parallelism within our experiments. Being able to adjust the task duration based on
    level of parallelism allowed us to achieve this.

@@

    Therefore, assuming a fixed total CPU time, the relation
    between batch and pilot jobs can be represented as:
    \begin{equation}
        \frac{M_{batch}}{M_{pilot}} = \frac{W_{pilot}}{W_{batch}}\label{eq:makespancomp}
    \end{equation}
    where:
    \begin{itemize}
        \item $M_{batch}$ is the makespan of the batch application
        \item $M_{pilot}$ is the makespan of the pilot application
        \item $W_{pilot}$ is the average number of workers of the pilot application
        \item $W_{batch}$ is the average number of workers of the batch application
    \end{itemize}
    We will use this relation to discuss our results later on. It
    corresponds to an ideal case where no data or other overhead is
    present: only queuing times are included, through the integration in
    Equation~\ref{eq:avgw}.

@@

As can be seen in both Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar}, pilots 
generally did not bring any performance improvement to the execution of Spark workloads
on B\'eluga and Cedar. In B\'eluga (Figure~\ref{fig:beluga1}) \tristan{why just in 1a and not 1b, ...?}, 16 pilots were always 
slower than batch except for repetitions 4 and 12. The maximum speedup achieved
was approximately 1.3~x for 8 pilots and 1.1~x for 16 pilots. In Figures~\ref{fig:beluga2}
and ~\ref{fig:beluga3}, it is also the case that batch is generally faster than
pilots for the majority of the cases, except a couple, in which the speedup brought
by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
that we see a maximum speedup of 2.8, however, it generally remains that
pilots are slightly slower. \TG{You should report the average speed-up of
pilots: 8 pilots vs batch and 16 pilots vs batch.}

@@

In order to determine whether queueing time was the cause of the difference 
in makespans between batch and pilots \tristan{rather than other overheads?}, we compared makespan with queuing time of 
first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
makespans is primiarily due to queueing time. \tristan{I don't like these figures: if you want to show
this, you should plot the makespan vs the queuing time of the first pilot, in an XY plot. But then the MW plot
is better because it takes into account the queuing time of all pilots, not only the first one. I would remove this figure.}

@@

In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
given an experiment configuration, where the line corresponds to the estimated makespan using 
our performance model. Makespan for the model was calculated as:
$$
M = \frac{125\times(d + 20)}{W}\times 10,
$$
where:
\begin{itemize}
    \item $125$ is the number of chunks BigBrain was split into
    \item $M$ is the makespan of the application, in seconds
    \item $W$ is the average number of Spark workers throughout the execution, computed as in Equation~\ref{eq:avgw}
    \item $d$ is the task delay associated with a given configuration
    \item $20$ is the average measured incrementation duration, in seconds, for a BigBrain chunk
    \item $10$ is the number of iterations
\end{itemize}
\tristan{You should remove the ceil from the plot too.}
As can be seen in the figure, although pilots and batch had similar makespans
for a given number of average workers, pilots were consistently slower than batch
for the same number of workers. Both batch and pilots were consistent with the 
model, however \TG{there is a contradiction here: if both batch and pilots were consistent with the model, then
they should have similar makespans.}. In general, both pilots and batch had the same number of average
workers. \TG{You could also plot the difference in the number of workers, for each experiment.}

@@

\tristan{Comment on the difficulties introduce by adding another layer in an already complex system:
HPC + Spark. It complexifies debugging.}
